Global code motion of congruent instructions

### 2017-04-18 paper.tex
use a few more gcm macros

\citet{click1995global} describes aggressive global code motion to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report sppeedup of as high as $23\%$.

@@

\section{Code motion}

The algorithm for code motion uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

@@

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}), and
\item update data structures to continue iterative code motion.
\end{itemize}

@@

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill), but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \gcm{} is less than before it will
be moved. Essentially, as long as there is one killed operand, code hoisting
will either decrease or preserve the register pressure.  Similarly, code-sinking
will either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how code motion of
identical computations can reduce the register pressure.

@@

\subsection{Code generation for the optimistic code motion algorithm}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting or sinking. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed, and the \SSA{} form is restored by
updating the intermediate representation (IR) to reflect the changes. At the
same time \MemorySSA{} is also updated to get up-to-date information about
memory references.

@@

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm as mentioned earlier in
Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
global code motion of congruent instructions which uses the liveness analysis as
illustrated by \citet{das2012} and ranking expressions explained by
\citet{rosen1988global}.

@@

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the global code motion of congruent computations in SSA
form. We saw that it improves inlining and LICM opportunities, reduces register
spills, it also improves performance in some cases. To preserve performance and
not hoist too much we have register pressure aware cost models described in
Section~\ref{subsec:cost-models}. A part of this implementation is already merged
in LLVM trunk as GVNHoist.cpp. We will post the \GCM{} implementation for review.

>>>>

\citet{click1995global} describes aggressive \gcm{} to first
schedule all the instructions as early as possible. This results in very long
live ranges which is mitigated by again scheduling all the instructions as late
as possible. They report sppeedup of as high as $23\%$.

@@

\section{Global code motion}

The algorithm for \gcm{} uses several common representations of the
program that we shortly describe below:
\begin{itemize}
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

@@

The \GCM{} pass can be broadly divided into the following steps that we will
describe in the rest of this section:
\begin{itemize}
\item find candidates (congruent instructions) suitable for \gcm{} (Section~\ref{subsec:finding-candidates}),
\item compute a point in the program where it is both legal
  (Section~\ref{subsec:legality}) and profitable
  (Section~\ref{subsec:cost-models}) to move the code,
\item move the code to hoist point or sink point
  (Section~\ref{subsec:optimistic}), and
\item update data structures to continue iterative \gcm{}.
\end{itemize}

@@

\subsubsection{Reduce register pressure}
\label{hoist:reg-pressure}
Hoisting upwards will decrease the live-range of its use, if it is a last use (a
kill), but increase the live-range of its definition. Conversely, sinking will
decrease the live-range of the defined register but increase the live-range for
killed operands. If the live-range after \gcm{} is less than before it will
be moved. Essentially, as long as there is one killed operand, code hoisting
will either decrease or preserve the register pressure.  Similarly, code-sinking
will either decrease or preserve the register pressure as long as there is one
operand killed at most.  The following example explains how \gcm{} of
identical computations can reduce the register pressure.

@@

\subsection{Code generation for the optimistic \gcm{} algorithm}
\label{subsec:optimistic}
Once all the legality and profitability checks are satisfied for a set of
congruent instructions, they are suitable candidates for hoisting or sinking. A
copy of the computation is inserted at the hoisting/sinking point along with any
instructions which needed to be rematerialized. Thereafter, all the computations
made redundant by the new copy are removed, and the \SSA{} form is restored by
updating the intermediate representation (IR) to reflect the changes. At the
same time \MemorySSA{} is also updated to get up-to-date information about
memory references.

@@

The amount of hoisting depends on whether we collect \GVN{} of instructions
before finding candidates (optimistic) or, on-demand (pessimistic). It also
depends on the generality of the \GVN{} algorithm as mentioned earlier in
Section~\ref{subsec:finding-candidates}. We have implemented an optimistic
\gcm{} of congruent instructions which uses the liveness analysis as
illustrated by \citet{das2012} and ranking expressions explained by
\citet{rosen1988global}.

@@

\section{Conclusion and Future Work}
\label{sec:future-work}
We have presented the \gcm{} of congruent computations in SSA
form. We saw that it improves inlining and LICM opportunities, reduces register
spills, it also improves performance in some cases. To preserve performance and
not hoist too much we have register pressure aware cost models described in
Section~\ref{subsec:cost-models}. A part of this implementation is already merged
in LLVM trunk as GVNHoist.cpp. We will post the \GCM{} implementation for review.
